---
output:
  word_document: default
  html_document: default
---
#CAPSTONE PROJECT – Hotel Bookings Cancellation Prediction   
    
####by:reena2930   
################################   
    
##Overview/Introduction    
Hotel bookings are always in high demand and there are many times we hope we get booking in our favourite hotel or resort. What is surprising to notice after I took up this project to the number of booking cancellations. This is defenitely bad news for hotel business as leads to wastage of food, resources, time and all leading to financial challenges.   
    
In this project I have aimed to develop an algorithm to predict booking cancellations based on the different features involved in the booking process. The data file has been downloaded from an existing available list of Kaggle datasets.   
    
ref: https://www.kaggle.com/jessemostipak/hotel-booking-demand   
     
This data set contains a single file which compares various booking information between two hotels: a city hotel and a resort hotel, and includes information such as when the booking was made, length of stay, the number of adults, children, and/or babies, and the number of available parking spaces, among other things.The csv file has been downloaded and made available as an additional attachment with the report. 
     
The dataset has been divided into 80-30 training and test sets and the final validation will be made on the test set. 


##Methods and Analysis   
The data set used is already in the clean form. I have attempted to format few columns, and remove few unwanted columns for easy analysis.
***PLEASE ATTACH THE SUPPORTING "hotel_bookingsCSV" file WHEN THE SYSTEM PROMPTS FOR THE CODING TO WORK***


###Dataset loading and creation of edx set and validation set     
      
          
# Note: This process could take a couple of minutes if you do not have the packages updated.          
```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(dslabs)) install.packages("dslabs", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
```
       
       
         
```{r}
#PLEASE ATTACH THE SUPPORTING hotel_bookingsCSV file WHEN THE SYSTME PROMPTS FOR THE CODING TO WORK.
hotel_bookings<-read.csv(file.choose(), header=T)
```
        
         
```{r}
head(hotel_bookings)
```
 
 
           
```{r}
#removing unwanted columns from the dataset
drop_cols = c("arrival_date_week_number", "is_repeated_guest", 
              "previous_bookings_not_canceled", "adults", "children", 
              "babies", "stays_in_weekend_nights", "stays_in_week_nights",
              "agent", "company", "days_in_waiting_list", "deposit_type", 
              "total_of_special_requests", "reservation_status", 
              "reservation_status_date", "country", 
              "required_car_parking_spaces")
```


```{r}
hoteldat<-hotel_bookings%>%select(-drop_cols)
```



```{r}
#dataset with revised number of columns
head(hoteldat)
```
     
     
     
    
```{r}
#using as_factor to convert variables into a factor to preserve the value and variable label attributes
hoteldat <- hoteldat %>%
  mutate(hotel = as_factor(hotel),
         is_canceled = as_factor(is_canceled),
         arrival_date_year = as_factor(arrival_date_year),
         arrival_date_month = as_factor(arrival_date_month),
         meal = as_factor(meal),
        market_segment = as_factor(market_segment),
        distribution_channel = as_factor(distribution_channel),
        previous_cancellations = as_factor(previous_cancellations),
        reserved_room_type = as_factor(reserved_room_type),
       assigned_room_type = as_factor(assigned_room_type)) 
```
  
  
     
```{r} 
#changing the 0 to No and 1 to Yes, for visualization purpose
hoteldat<-hoteldat%>%
  mutate(is_canceled = ifelse(str_detect(is_canceled,"0")==TRUE,"No","Yes"))
```



```{r}
summary(hoteldat)
```

  
  
   
```{r}
head(hoteldat)
```




```{r}
dim(hoteldat)
```
      

The cleaned data now has 119390 rows and 15 columns.   
Both the structure and head options show no missing values and NA's. We can now proceed with splitting the data into training and test sets, in the proportion of 80-20   


```{r}         
set.seed(1, sample.kind="Rounding")
         #we will use a data partition set with 80% training data and 20% test data
test_index <- createDataPartition(y = hoteldat$is_canceled, times = 1, p = 0.2, list = FALSE)
train_set<-hoteldat[-test_index,]
test_set<-hoteldat[test_index,]
```
 
 
 
##**Data Exploration**       
```{r}
summary(train_set)
```

The summary of the subset shows that the edx set has 95511 observations with 15 variables and there are no missing values or NA. Each feature represents individual column in the dataset. The test set has the same features except that its 20% of the total dataset.



```{r}         
train_set%>%group_by(is_canceled)%>%summarize(n=n())
```
  
 
     
```{r}
head(train_set)
```

Above is a glimpse of how the data looks. Each row represents data realted to the hotel booking. The “is_canceled” column is the outcome(y) we want to predict. Let us now look at the basic features and characteristics of the datasets




###Proportion of bookings that were canceled with both Hotels  

City Hotel seems to have more cancellations than the Resort hotel   
```{r}
#Proportion of bookings that were canceled
train_set%>%ggplot(aes(x=hotel, fill=is_canceled))+geom_bar()+ggtitle("Canceled Booking by Hotel")+geom_text(stat = "count", aes(label=..count..),vjust=-0.1)
```




###Bookings canceled between 2015-2017  
```{r}
train_set%>%ggplot(aes(x=arrival_date_year, fill=is_canceled))+geom_bar()+ggtitle("Canceled Bookings Trend by Years")+geom_text(stat = "count", aes(label=..count..),vjust=-0.05)
```

Cancellations were high in 2016 and reduced in 2017, which is still higher than 2015. 2015 shows the lowest cancellations off the 3 years.   


     
###Monthly Booking cancellations    
```{r}        
train_set%>%ggplot(aes(x=arrival_date_month, fill=is_canceled))+geom_bar()+coord_flip()+ggtitle("Canceled Booking by Months")+geom_text(stat = "count", aes(label=..count..),vjust=-0.05)
```
August shows the highest cancellations followed by April, July, June and May. Hotel administration will need to look into the reasons behind this peak in Aug and make necessary changes to boost non cancellations




###Bookings canceled based lead time  
Lead Time is the period of time between when a guest makes a reservation, and the actual check-in date.   
```{r}         
train_set%>%ggplot(aes(x=lead_time, fill=is_canceled))+geom_bar()+ggtitle("Canceled Booking by Lead Time")
``` 
There are not many cancellations after the lead time has elapsed. cancellations soon after booking seem to be a trend.   



###Bookings canceled based Market segment   
Most of the time bookings via different market segments bring in more business then direct hotel bookings. Let us look at what the trend shows here
```{r}         
train_set%>%ggplot(aes(x=market_segment, fill=is_canceled))+geom_bar()+coord_flip()+ggtitle("Canceled Booking by Market Segment")
```
This definitely needs to be looked into by the Hotel administration. Bookings ,made by TA's and TO's via online medium show highest cancellations, though their bookings are good. Group bookings, which bring in more revenue, are second highest in cancellations.



###Bookings canceled based on Previous Cancellations   

```{r}         
train_set%>%ggplot(aes(x=previous_cancellations, fill=is_canceled))+geom_bar()+coord_flip()+ggtitle("Canceled Bookings by Previous Cancellations")
```
Cancellations in rationship to previous cancellations are steep, which is obvious considering the history.   






##**Model Preparation**


###LDA model
A relatively simple solution to the problem of having too many parameters is to assume that the correlation structure is the same for all classes, which reduces the number of parameters we need to estimate. We can fit the LDA model using caret.
One model is based only using train_set and the second model is used on the test_set

```{r}
#LDA model with train set
set.seed(1, sample.kind = "Rounding")

train_lda <- train(is_canceled ~ arrival_date_year + lead_time + adr + arrival_date_month + booking_changes, method = "lda", data = train_set)
lda_preds <- predict(train_lda, train_set)
confusionMatrix(data=lda_preds, reference=factor(train_set$is_canceled))$overall["Accuracy"]
```

```{r}
#LDA model with test set
test_lda <- train(is_canceled ~ arrival_date_year + lead_time + adr + arrival_date_month + booking_changes, method = "lda", data = train_set)
lda_preds <- predict(test_lda, test_set)
confusionMatrix(data=lda_preds, reference=factor(test_set$is_canceled))$overall["Accuracy"]
```


**LDA method**  | **Accuracy**
--------------- | -------------
Training Set    |  0.6815131
Test Set        |  0.6809749



##**Logistic regression model**    

The simplest prediction method is randomly guessing the outcome without using additional predictors. These methods will help us determine whether our machine learning algorithm performs better than chance.

test_glm<-train(is_canceled~., method="glm", data=test_set)


We will use the glm method with few variations, with both train set and test set

##**1st glm model**    
```{r}
#1st glm model on train set
set.seed(1, sample.kind = "Rounding")
train_glm<- train(is_canceled ~ arrival_date_year + lead_time + adr, method = "glm", data = train_set)
glm_pred <- predict(train_glm, train_set)
confusionMatrix(data=glm_pred, reference=factor(train_set$is_canceled))$overall["Accuracy"]
```


```{r}
#1st glm model on test set
test_glm<- train(is_canceled ~ arrival_date_year + lead_time + adr, 
method = "glm", data = train_set)
glm_pred <- predict(test_glm, test_set)
confusionMatrix(data=glm_pred, reference=factor(test_set$is_canceled))$overall["Accuracy"]
```



GLM Method with 3 predictors   
**GLM method**  | **Accuracy**
--------------- | -------------
Training Set    |  0.6641643
Test Set        |  0.6645588 



###**2nd glm model**    
```{r}
#2nd glm model on train set
set.seed(1, sample.kind = "Rounding")
train_glm2<- train(is_canceled ~ arrival_date_year + lead_time + adr + arrival_date_month + booking_changes, method = "glm", data = train_set)
glm_pred2 <- predict(train_glm2, train_set)
confusionMatrix(data=glm_pred2, reference=factor(train_set$is_canceled))$overall["Accuracy"]
```


```{r}
#2nd glm model on test set
test_glm2<- train(is_canceled ~ arrival_date_year + lead_time + adr + arrival_date_month + booking_changes, method = "glm", data = train_set)
glm_pred2<- predict(test_glm2, test_set)
confusionMatrix(data=glm_pred2, reference=factor(test_set$is_canceled))$overall["Accuracy"]
```

GLM Method with 5 predictors   
**GLM method2**  | **Accuracy**
--------------- | -------------
Training Set    |  0.6645588
Test Set        |  0.6837389   



##**RESULT**    

The model is able to predict cancellations with an accuracy of 68%. The accuracy levels with both LDA and glm were 68% but glm method came with warning notes. LDA model fits our requirement for the requirement of this project. 

**LDA method**  | **Accuracy**
--------------- | -------------
Training Set    |  0.6815131
Test Set        |  0.6809749

The bookings and cancellations were on higher side on the City Hotel compared to Resort Hotel and Online bookings by TA’s and Group cancellations, combined by the tradition of previous cancellation trend and lead time contributed towards cancellation more than other predictors in the data set.  


##**CONCLUSION**     

It can be concluded that our recommended LDA model is able to predict the cancellation of hotel booking by 68% considering the few predictors used. However, the data can be formatted further and similar columns combined, like family numbers, booking segments etc., and more features explored to get a better accuracy level. We also can look at more advanced regression techniques to achieve better levels of accuracy.



******End******* 